Only the store can save you now:\\updates and vulnerabilities on Android

### 2015-04-09 securityupdates.tex
Minor fixes on reading through

% Hypothesis
% Attempts at fine grained restrictions on running arbitrary code are hampered by updates for security vulnerabilities not reaching users in a timely fashion.
\vspace{-1em}
\begin{abstract}
Modern smartphone operating systems protect users from malicious applications by providing a protected execution environment or sandbox.
A number of exploits have been found which allow apps to break out of the sandbox.
We quantify for the first time how vulnerable Android devices in the wild are to critical vulnerabilities and how long it takes for vulnerabilities to be fixed on consumer devices.
We analysed \da\ data from over \daOSYearsOfData\ years and \daNumOSDevices\ devices and found that on average \daMeanInsecurityPercTwosfNominal\ of devices were exposed to known root privilege vulnerabilities and only \daUpdatednessPercTwosfNominal\ of devices run the most recent version of Android.
Devices apply \daUpdatesPerYearTwosfNominal\ updates each year, less than the rate of critical vulnerability discovery of between \avoVulnsPerYearAllAndroidTwosfNominal\ and \avoVulnsPerYearTwosfNominal\ per year.
\end{abstract}

\section{Introduction}
\todo{We talk about root a lot again, rather critical vulnerabilities, and root is not strictly accurate}
Modern mobile operating systems provide several layers of security whose combined aim is to prevent app malware from taking control of a device. 
In this paper we explore the security layers present in the Android ecosystem and determine which layers are effective and which are not.

Our focus is on malware whose aim is to gain full control of the device rather than merely abuse permissions granted to it by the user.
In this context, by full control, or \emph{root}, on a device we mean the malware has unfettered access to all data stored on the device and can control all network interfaces in the device with the same flexibility as the operating system itself.
We have selected our focus for two reasons: no previous published work has quantified how vulnerable Android devices are to malware gaining root, and secondly the negative impact on the device owner is much higher for malware which gains root than malware which does not: as it can not only perform arbitrary actions on the device, but it is also impossible for the device owner to remove the malware reliably as even a factory reset does not guarantee that the malware has been removed.

@@

Android users are protected from malicious apps by up to three levels of defence. 
The first level of defence comes from an app marketplace such as Google Play. 
A marketplace can use three mechanisms to prevent malicious apps from appearing in the marketplace: firstly by economic disincentives, including charging developers to create an account and cancelling malicious accounts; secondly by technical means, such as static and dynamic analysis on submitted apps; and finally via social feedback from users, including the provision of reviews, rating and reporting. 
The second line of defence occurs on the handset at installation time, and includes checking the signature of any app installation or update as well as statically verifying the app (e.g. Google's Verify Apps feature).
Finally, the third line of defence takes place at runtime: Android apps are run in a \emph{sandbox} which prevents the app from unauthorised actions such as accessing any permissions which weren't granted by the user at install time; in addition, anti-virus products may scan the list of installed apps periodically to detect malware.

@@

In summary, our contributions are:
\begin{itemize}
 \item We quantify the Android update ecosystem showing how the OS version distribution changes and the frequency of updates.
 \item We created an open database to collate quantitative information on vulnerabilities which affect Android and use this to determine the proportion of devices they affect and how that changes over time.
 \item We demonstrate that the security of Android relies on the centralisation of app deployment via a marketplace, and the use of install-time anti-virus checks rather than through the security of the app sandbox.
 \item We use our analysis to develop advice for Android users on what steps they should take to increase the security of their device.
 \item We characterise the impact of critical vulnerabilities on the Android ecosystem as a whole, examine the effectiveness of vulnerabilities and how long vulnerabilities remained available in the wild, and also examine the impact on individual devices over extended periods of time.

@@

We use two sources of data to determine whether the Android sandbox is secure or not: (1) information on the distribution of installed versions of Android over time and (2) information on the critical vulnerabilities found to affect specific versions of Android.
These two datasets can then be combined to determine the proportion of sandboxes at risk of attack from specific vulnerabilities over time.

@@

The \da\ app collects a range of metadata from Android devices.\footnote{\url{https://deviceanalyzer.cl.cam.ac.uk/collected.html}}
We extracted the build string and API version for each device each day.
The build string is a user-readable version string.
The API version is a positive integer which increases when new features are added to the API.
Consequently security (bug) fixes do not result in a change in the API version.
Fortunately most (\daOSVersionPercValidLines) entries in the data have a build string of the form `x.y.z opaque\_marker' and so it is possible to extract the Android version number `x.y.z'.
On a large proportion of devices `opaque\_marker' is a well defined build number\footnote{\url{https://source.android.com/source/build-numbers.html}} however different device manufacturers use different schema.
Google provides API version distribution information but not the OS and build version information we need. We verify that the \da\ data is representative in~\S\ref{sec:representative}.

@@

%The discovery dates of our vulnerabilities are not uniform. 
%In particular the data shows a large gap from 2011-10-06 to 2013-02-18 where we do not know of any discoveries of critical vulnerabilities affecting all Android devices.
%The cause of this quiet period is unclear.
%Possible explanations are that: (i) most devices were exposed to known vulnerabilities so there was reason to look for new ones;
%(ii) device manufacturers made it easier to install custom versions of Android, reducing the need for users to root their devices; and
%(iii) device manufacturer specific vulnerabilities (which are not in our analysis here) were easier to find and so attackers looking for vulnerabilities adjusted their focus.

@@

\subsubsection{Method} 
To investigate the vulnerability of Android devices, we used the OS version information from \da\ and the vulnerability data from \avo.
The \da\ data was collected between \daStartDate\ and \daEndDate.
The \avo\ data covers the period from \avoFirstDataDate\ to \avoLastDataDate.
For each device we collected daily version data and any vulnerabilities it was exposed to at that time.
We then normalise these totals for each day by dividing through by the total number of devices with version information on that day.

@@

Following time from left to right in Figure~\ref{fig:proportioninsecure} we see all devices are initially \emph{maybe secure} (yellow) since \da\ does not have historical data prior to May 2011. This means we cannot distinguish between devices which are running a version of Android which is known to be vulnerable from one which may have received a backported fix.
This demonstrates the importance of a longitudinal study: this type of analysis requires years of data.
Once \avovuln{zergRush} was discovered in October 2011 then most devices are recorded as \emph{insecure} (red) as they were exposed to that vulnerability.
The remaining devices were already running a version of Android which fixed the \avovuln{zergRush} vulnerability and are therefore marked as \emph{secure} (green).
From October 2011 until the discovery of \avovuln[APK_duplicate_file]{APK duplicate file} in February 2013 the graph shows progressive improvement as devices are upgraded.
This means more and more devices are marked as \emph{secure} because they are now running a secure version of Android, or marked as \emph{maybe secure} because they received an OS update which did not update to a known-good version of Android but which may still have included a backport of a fix, as the update was made available after the vulnerability was disclosed.
From February 2013 onwards regular discovery of critical vulnerabilities ensures that most devices are exposed to known critical vulnerabilities.

This gives us an average exposure to critical vulnerabilities of Android devices of \daMeanInsecurityPerc.

@@

In 2013 three vulnerabilities were found in the way which Android verified the signatures on APKs.
These are level 2 vulnerabilities in our threat model since they require a new app installation to occur.
Figure~\ref{fig:nvulnerabilities_heat} shows how the the \emph{APK signing vulnerabilities} affected all devices and took months to get fixed for any device.
However what is perhaps more worrying is the long tail on the \avovuln{Gingerbreak}, \avovuln{levitator}, \avovuln[exploid_udev]{exploid} and \avovuln{zergRush} vulnerabilities which are more dangerous level 3 vulnerabilities (since they do not require new app installation) and which affect a significant proportion of devices years later.
%\todo{quantify this long tail and the APK numbers as headline numbers}

@@

{
\begin{figure}[!t]
\centering
\begin{subfigure}{\columnwidth}
\includegraphics[width=\columnwidth]{figures/device-data-all-security}
\vspace{-1.5em}
\caption{Each strip shows the number of vulnerabilities affecting each device over time.}
\label{fig:device_data_security}
\end{subfigure}
\begin{subfigure}{\columnwidth}
\includegraphics[width=\columnwidth]{figures/device-data-all-os}
\vspace{-1.5em}
\caption{Each strip shows OS versions over time. Black lines show where an update only changed the build number.}
\label{fig:device_data_os}
\end{subfigure}
\caption{The top \daNumDeviceDataDevices\ devices by days of contribution in the \da\ data. One strip per device handset.}
\end{figure}
}

@@

Currently, finding and exploiting vulnerabilities is typically a time-consuming manual process.
Therefore an attacker has to invest significant human resource to first find a bug and then write code to exploit the vulnerability.
To reduce the effort involved in finding a bug, attackers can look at security-related commits made to open source repositories.
Attackers have taken this approach for the Firefox web browser~\cite{Barth2011}.
Google does not release updates to the Android source code until after the release of the relevant update.
This reduces the time available for an attacker to exploit a vulnerability by looking at security updates disclosed in the source code repository.
Reducing it from, the interval between the commit of the first fix and the fix being deployed to most devices, to the interval between the first fixing release and the fix being deployed to most devices; unfortunately this can still take months or years.
For example, the zergRush vulnerability was fixed in a release before it was publicly discovered but it still took 27 months for 90\% of the devices in our study to be definitely fixed\footnote{Definitely fixed is 1 $-$ (definitely insecure + maybe secure), maybe fixed is 1 $-$ definitely insecure} (25 months for 90\% of devices maybe fixed).

Brumley et al. has shown it is possible to automatically generate exploits from binary fixes, for example by using security updates issued by manufacturers~\cite{Brumley2008}. 
Similarly, the Woodpecker tool automatically finds permission leaks in stock Android phone images~\cite{Grace2012}.
Once such automated techniques mature, it will become even more important to provide security updates promptly to every Android device in the ecosystem because an attacker can turn the contents of a released fix into an exploit in a matter of minutes or hours.

@@

There are continuing efforts to reduce the impact of critical vulnerabilities, both in Android and elsewhere.
SEAndroid~\cite{Smalley2013} which is included in Android from version 4.1~\cite{jelly-bean-release} claimed to prevent some root vulnerabilities and to reduce the impact of others.
Capability based enforcement systems such as Capsicum~\cite{Watson2010} substantially reduce the capabilities that an exploit has to try and gain increased privilege with and could be included in Linux\footnote{\url{https://github.com/google/capsicum-linux}} and hence Android.

@@

Unfortunately, whilst these tools were successful in finding critical vulnerabilities, such techniques are not perfect.
The Malware Genome project collected 1\,260 malware samples from 2010--2011~\cite{Zhou2012b} and found a best case anti-malware detection rate of 79.6\%.
Similarly, DroidChameleon found that antivirus products could not detect malware if the binary was automatically permuted~\cite{Rastogi2013}.
Therefore detecting risky apps which might contain critical vulnerabilities is a useful tool, but it does not replace the need to provide security updates.

@@

Installing regular updates is the best defence against malware, however, as we have shown, these updates are rarely forthcoming. 
Therefore Android users need to take additional steps to protect themselves. 
Firstly, users should only install apps from Google Play, which previous research has been shown to have a lower level of malware than third-party markets and is known to contain static and dynamic analysis of available apps.
Secondly, users should consider installing an anti-virus tool or enabling the Google verify apps feature if available. 
Thirdly, users should avoid using untrusted WiFi networks or use them in conjunction with a VPN tunnel to a trusted network. 
Finally, if the user is running a device with API level 17 or above and is technically able to determine the target API version of their apps, they should consider uninstalling all apps with a target API of less than 17.

>>>>

% Hypothesis
% Attempts at fine grained restrictions on running arbitrary code are hampered by updates for security vulnerabilities not reaching users in a timely fashion.
\vspace{-1em}
\begin{abstract}
Modern smartphone operating systems protect users from malicious applications by providing a protected execution environment or sandbox.
A number of exploits have been found which allow apps to break out of the sandbox.
We quantify for the first time how vulnerable Android devices in the wild are to critical vulnerabilities and how long it takes for vulnerabilities to be fixed on consumer devices.
We analysed \da\ data from over \daOSYearsOfData\ years and \daNumOSDevices\ devices and found that on average \daMeanInsecurityPercTwosfNominal\ of devices were exposed to known root privilege vulnerabilities and only \daUpdatednessPercTwosfNominal\ of devices run the most recent version of Android.
Devices apply \daUpdatesPerYearTwosfNominal\ updates each year, less than the rate of critical vulnerability discovery of between \avoVulnsPerYearAllAndroidTwosfNominal\ and \avoVulnsPerYearTwosfNominal\ per year.
\end{abstract}
\todo{Abstract does not reflect current spin very well}

\section{Introduction}
Modern mobile operating systems provide several layers of security whose combined aim is to prevent app malware from taking control of a device. 
In this paper we explore the security layers present in the Android ecosystem and determine which layers are effective and which are not.

Our focus is on vulnerabilities which allow malware to gain full control of the device rather than merely abuse permissions granted to it by the user.
In this context, by full control, or \emph{root}, on a device we mean the malware has unfettered access to all data stored on the device and can control all network interfaces in the device with the same flexibility as the operating system itself.
We have selected our focus for two reasons: no previous published work has quantified how vulnerable Android devices are to malware gaining root, and secondly the negative impact on the device owner is much higher for malware which gains root than malware which does not: as it can not only perform arbitrary actions on the device, but it is also impossible for the device owner to remove the malware reliably as even a factory reset does not guarantee that the malware has been removed.

@@

Android users are protected from malicious apps by up to three levels of defence. 
The first level of defence comes from an app marketplace such as Google Play. 
A marketplace can use three mechanisms to prevent malicious apps from appearing in the marketplace: firstly by economic disincentives, including charging developers to create an account and cancelling malicious accounts; secondly by technical means, such as static and dynamic analysis on submitted apps; and finally via social feedback from users, including the provision of reviews, rating and reporting. 
The second line of defence occurs on the handset at installation time, and includes checking the signature of any app installation or update as well as checking app for malware (e.g. Google's Verify Apps feature).
Finally, the third line of defence takes place at runtime: Android apps are run in a \emph{sandbox} which prevents the app from unauthorised actions such as accessing any permissions which weren't granted by the user at install time; in addition, anti-malware products may scan the list of installed apps periodically to detect malware.

@@

In summary, our contributions are:
\begin{itemize}
 \item We quantify the Android update ecosystem showing how the OS version distribution changes and the frequency of updates.
 \item We created an open database to collate quantitative information on vulnerabilities which affect Android and use this to determine the proportion of devices they affect and how that changes over time.
 \item We demonstrate that the security of Android relies on the centralisation of app deployment via a marketplace, and the use of install-time anti-malware checks rather than through the security of the app sandbox.
 \item We use our analysis to develop advice for Android users on what steps they should take to increase the security of their device.
 \item We characterise the impact of critical vulnerabilities on the Android ecosystem as a whole, examine the effectiveness of vulnerabilities and how long vulnerabilities remained available in the wild, and also examine the impact on individual devices over extended periods of time.

@@

We use two sources of data to determine whether the Android sandbox is secure or not: (1) information on the distribution of installed versions of Android over time and (2) information on the critical vulnerabilities found to affect specific versions of Android.
These two datasets can then be combined to determine the proportion of devices at risk of attack from specific vulnerabilities over time.

@@

The \da\ app collects a range of data from Android devices.\footnote{\url{https://deviceanalyzer.cl.cam.ac.uk/collected.html}}
We extracted the build string and API version for each device each day.
The build string is a user-readable version string.
The API version is a positive integer which increases when new features are added to the API.
Consequently security (bug) fixes do not result in a change in the API version.
Fortunately most (\daOSVersionPercValidLines) entries in these data have a build string of the form `x.y.z opaque\_marker' and so it is possible to extract the Android version number `x.y.z'.
On a large proportion of devices `opaque\_marker' is a well defined build number\footnote{\url{https://source.android.com/source/build-numbers.html}} however different device manufacturers use different schema.
Google provides API version distribution information but not the OS and build version information we need. We verify that the \da\ data is representative in~\S\ref{sec:representative}.

@@

%The discovery dates of our vulnerabilities are not uniform. 
%In particular these data shows a large gap from 2011-10-06 to 2013-02-18 where we do not know of any discoveries of critical vulnerabilities affecting all Android devices.
%The cause of this quiet period is unclear.
%Possible explanations are that: (i) most devices were exposed to known vulnerabilities so there was reason to look for new ones;
%(ii) device manufacturers made it easier to install custom versions of Android, reducing the need for users to root their devices; and
%(iii) device manufacturer specific vulnerabilities (which are not in our analysis here) were easier to find and so attackers looking for vulnerabilities adjusted their focus.

@@

\subsubsection{Method} 
To investigate the vulnerability of Android devices, we used the OS version information from \da\ and the vulnerability data from \avo.
The \avo\ data covers the period from \avoFirstDataDate\ to \avoLastDataDate.
The \da\ data was collected between \daStartDate\ and \daEndDate.
For each device we collected daily version data and any vulnerabilities it was exposed to at that time.
We then normalise these totals for each day by dividing through by the total number of devices with version information on that day.

@@

Following time from left to right in Figure~\ref{fig:proportioninsecure} we see all devices are initially \emph{maybe secure} (yellow) since \da\ does not have historical data prior to May 2011. This means we cannot distinguish between devices which are running a version of Android which is known to be vulnerable from one which may have received a backported fix.
This demonstrates the importance of a longitudinal study: this type of analysis requires years of data.
Once \avovuln{zergRush} was discovered in October 2011 then most devices are recorded as \emph{insecure} (red) as they were exposed to that vulnerability.
The remaining devices were already running a version of Android which fixed the \avovuln{zergRush} vulnerability and are therefore marked as \emph{secure} (green).
From October 2011 until the discovery of \avovuln[APK_duplicate_file]{APK duplicate file} in February 2013 the graph shows progressive improvement as devices are upgraded.
This means more and more devices are marked as \emph{secure} because they are now running a secure version of Android, or marked as \emph{maybe secure} because they received an OS update which did not update to a known-good version of Android but which may still have included a backport of a fix, as the update was made available after the vulnerability was disclosed.
From February 2013 onwards regular discovery of critical vulnerabilities ensures that most devices are exposed to known critical vulnerabilities.
This gives us an average exposure to critical vulnerabilities of Android devices of \daMeanInsecurityPerc.

@@

In 2013 three vulnerabilities were found in the way which Android verified the signatures on APKs.
These are level 2 vulnerabilities in our threat model since they require a new app installation to occur.
Figure~\ref{fig:nvulnerabilities_heat} shows how the the \emph{APK signing vulnerabilities} affected all devices and took months to get fixed for any device.
However what is perhaps more worrying is the long tail on the \avovuln{Gingerbreak}, \avovuln{levitator}, \avovuln[exploid_udev]{exploid} and \avovuln{zergRush} vulnerabilities which are more dangerous level 3 vulnerabilities (since they do not require new app installation) and which affected a significant proportion of devices years later.
%\todo{quantify this long tail and the APK numbers as headline numbers}

@@

{
\begin{figure}[!t]
\centering
\begin{subfigure}{\columnwidth}
\includegraphics[width=\columnwidth]{figures/device-data-all-security}
\vspace{-1.5em}
\caption{Each strip shows the number of vulnerabilities affecting each device over time.}
\label{fig:device_data_security}
\end{subfigure}
\begin{subfigure}{\columnwidth}
\includegraphics[width=\columnwidth]{figures/device-data-all-os}
\vspace{-1.5em}
\caption{Each strip shows OS versions over time.}
\label{fig:device_data_os}
\end{subfigure}
\caption{The top \daNumDeviceDataDevices\ devices by days of contribution in the \da\ data. One strip per device handset. Black lines show where an update only changed the build number.}
\end{figure}
}

@@

Currently, finding and exploiting vulnerabilities is typically a time-consuming manual process.
Therefore an attacker has to invest significant human resource to first find a bug and then write code to exploit the vulnerability.
To reduce the effort involved in finding a bug, attackers can look at security-related commits made to open source repositories.
Attackers have taken this approach for the Firefox web browser~\cite{Barth2011}.
Google does not release updates to the Android source code until after the release of the relevant update~\cite{AndroidSecurity2014}.
This reduces the time available for an attacker to exploit a vulnerability by looking at security updates disclosed in the source code repository.
Reducing it from, the interval between the commit of the first fix and the fix being deployed to most devices, to the interval between the first fixing release and the fix being deployed to most devices; unfortunately this can still take months or years.
For example, the zergRush vulnerability was fixed in a release before it was publicly discovered but it still took 27 months for 90\% of the devices in our study to be definitely fixed\footnote{Definitely fixed is 1 $-$ (definitely insecure + maybe secure), maybe fixed is 1 $-$ definitely insecure} (25 months for 90\% of devices to be maybe fixed).

Brumley et.\ al. has shown it is possible to automatically generate exploits from binary fixes, for example by using security updates issued by manufacturers~\cite{Brumley2008}. 
Similarly, the Woodpecker tool automatically finds permission leaks in stock Android phone images~\cite{Grace2012}.
Once such automated techniques mature, it will become even more important to provide security updates promptly to every Android device in the ecosystem because an attacker can turn the contents of a released fix into an exploit in a matter of minutes or hours.

@@

There are continuing efforts to reduce the impact of critical vulnerabilities, both in Android and elsewhere.
SEAndroid~\cite{Smalley2013} which is included in Android from version 4.1~\cite{jelly-bean-release} and fully enforcing from version 5.0~\cite{AndroidSecurity2014} claimed to prevent some root vulnerabilities and to reduce the impact of others.
Capability based enforcement systems such as Capsicum~\cite{Watson2010} substantially reduce the capabilities that an exploit has to try and gain increased privilege with and could be included in Linux\footnote{\url{https://github.com/google/capsicum-linux}} and hence Android.

@@

Unfortunately, whilst these tools were successful in finding malware, such techniques are not perfect.
The Malware Genome project collected 1\,260 malware samples from 2010--2011~\cite{Zhou2012b} and found a best case anti-malware detection rate of 79.6\%.
Similarly, DroidChameleon found that antivirus products could not detect malware if the binary was automatically permuted~\cite{Rastogi2013}.
Therefore detecting risky apps which might contain critical vulnerabilities is a useful tool, but it does not replace the need to provide security updates.

@@

Installing regular updates is the best defence against malware, however, as we have shown, these updates are rarely forthcoming. 
Therefore Android users need to take additional steps to protect themselves. 
Firstly, users should only install apps from Google Play, which previous research has been shown to have a lower level of malware than third-party markets and is known to contain static and dynamic analysis of available apps.
Secondly, users should enable the Google Verify Apps feature if available or use an anti-malware app if not. 
Thirdly, users should avoid using untrusted WiFi networks or use them in conjunction with a VPN tunnel to a trusted network. 
Finally, if the user is running a device with API level 17 or above and is technically able to determine the target API version of their apps, they should consider uninstalling all apps with a target API of less than 17.
